|

### 2021-12-08 submission/manuscript.rmd
incorporate comments and update rmse figures

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting:** We simulated data using diverse assumptions for a baseline
 prognostic index of risk (PI), the shape of its interaction with
 treatment (none, linear, quadratic or non-monotonic) and the magnitude
 of treatment-related harms. In each sample we predicted absolute
 benefit using: models with a constant relative treatment effect;
 stratification in quarters of the PI; models including a linear
 interaction of treatment with the PI; models including an interaction
 of treatment with a restricted cubic spline (RCS) transformation of
 the PI; an adaptive approach using Akaike’s Information Criterion. We
 evaluated predictive performance using root mean squared error and
 measures of discrimination and calibration for benefit. Starting from
 a base scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
 the PI 0.75), we varied the sample size, the treatment effect size,
 and the PI's AUC. **Results:** Models including a PI by treatment
 interaction performed best with smaller sample sizes and/or lower AUC
 of the PI. Otherwise, RCS (3 knots) models proved more robust for the
 majority of the deviation settings. The adaptive approach was unstable
 with smaller sample sizes. **Conclusion:** Depending on the setting,
 the linear interaction or the RCS (3 knots) model should be preferred.

authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. However, even though estimates at the risk subgroup level may be
accurate, these estimates do not apply to individual patients, especially for
patients with predicted risk at the boundaries of the risk intervals. Hence, the
risk-stratified approach is useful for exploring and presenting HTE, but is not
useful for supporting treatment decisions for individual patients.

@@

For each patient we generated 8 baseline covariates
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
was allocated using a 50:50 split. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates. In
the base scenarios coefficient values were such, that the AUC of the
logistic regression model was $0.75$ and the event rate in the control arm was
$20\%$. Binary outcomes in the control arm were generated from Bernoulli variables
with true probabilities 
$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.

Outcomes in the treatment arm were generated using 3 base scenarios: absent
treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and high
treatment effect (OR = 0.5). We started with simulating outcomes based on true
constant relative treatment effects for the 3 base scenarios. We then simulated
linear, quadratic and non-monotonic deviations from constant treatment effects
using:
$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
constant absolute harm is applied across all treated patients. In this case we
have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,063 and 17,000, respectively. We also evaluated the effect of
worse or better discriminative ability for risk, adjusting the baseline
covariate coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 648 scenarios
(exact settings in the supplementary material).

@@

All methods assume that a risk prediction model is available to assign risk
predictions to individual patients. For the simulations we developed a
prediction model internally, using logistic regression including main effects
for all baseline covariates and treatment assignment. Risk predictions for
individual patients were based on treatment assignment to the control arm, that
is setting treatment assignment to 0.

@@

```{r adaptive, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveSelections.csv"))
rmseDistribution <- readr::read_csv(here::here("data/processed/rmseDistribution.csv"))
aucDistribution <- readr::read_csv(here::here("data/processed/discriminationDistribution.csv"))
```
The linear interaction model outperformed all RCS methods in terms of
RMSE in scenarios with true constant relative treatment effect (OR =
0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
quadratic deviations (Figure \ref{fig:rmsebase}). However, with
non-monotonic deviations errors of the linear interaction model
increased substantially, especially in the presence of
treatment-related harms. In these scenarios, RCS-3 proved to be quite
robust outperforming all other methods. The constant treatment effect
approach had overall best performance under true constant treatment
effect settings, but was sensitive to all considered deviations,
resulting in increased error rates. Finally, the adaptive approach had
comparable performance to the best-performing method in each
scenario. However, we observed increased error variability in the case
of linear and non-monotonic deviations, especially for moderate or
strong treatment-related harms.

In the case of true constant relative treatment effects, the adaptive
approach selected the constnant effect model the majority of the time,
even in the presence of treatment-related harms
(`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\%
of the time with no harms to
`r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
with strong harms). With strong linear or quadratic deviations and
increasing treatment-related harms the adaptive approach increasingly
favored the linear interaction model. In the case of true
non-monotonic deviations, we observed increasing selection frequencies
for RCS-3 with increasing treatment harms (from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
with strong harms), whereas selection frequencies for the linear
interaction model dropped from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
with strong harms. Finally, in the case of strong linear and
non-monotonic deviations selection frequencies of the constant
treatment effect model remained high despite increasing treatment
harms (Supplement, Table XX).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000. The scenario with true constant relative treatment effect had a true prediction AUC of 0.75 and sample size of 4250. Results are presented under moderate linear, strong linear, and strong quadratic deviations from constant relative treatment effects.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

Increasing the sample size to `r 17000` favored RCS-3 the most, as it
achieved lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower error rates (median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of treatment-related harms strength. The
issues with the large error variability of the adaptive approach
improved with larger sample sizes.

The adaptive approach tended to increasingly favor smoother methods
(especially RCS-3) with increasing treatment-related harms (see
Supplement, Table XX). However, in the case of true strong quadratic
deviations the opposite was observed: selection frequency of the
linear interaction model increased from
`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(risk) * 100`\% (no harms)
to
`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(risk) * 100`\% (strong harms)
whereas for RCS-3 decreased from
`r adaptiveSelections %>% filter(scenarioId == 385) %>% pull(rcs_3) * 100`\% (no harms)
to
`r adaptiveSelections %>% filter(scenarioId == 387) %>% pull(rcs_3) * 100`\% (strong harms).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

When we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = `r 4250`) the constant effect model outperformed all other
methods in the case of true constant treatment effects, but proved to
be the least robust to deviations. Again, RCS-3 had the lowest
error rates in the case of strong quadratic or monotonic deviations
and very comparable performance to the best-performing linear
interaction model in the case of strong linear deviations
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). The adaptive approach, though it
performed similar to the best performing method in each scenario, on
average, had increased variability in error rates in the case of
strong linear and non-monotonic deviations. In these scenarios, the
adaptive approach often selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
in the strong linear and non-monotonic deviation scenarios without
treatment-related harms, respectively).

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
= 0.75), all methods performed similarly, on average (Figure
\ref{fig:discrimination}). In the case of non-monotonic deviations,
the constant effect model had much lower discriminative performance
compared to the rest of the methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model compared to the best-performing RCS-3
with
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
The linear interaction model was the most stable compared to the other
methods in terms of error variability. With increasing number of RCS
knots, we observed decreasing median values and increasing variability
of the c-for-benefit in all scenarios.

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the case of true constant treatment
effects, but was miscalibrated for all deviation scenarios (Figure
\ref{fig:calibration}). The linear interaction model showed best or
close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in the case of non-monotonic deviations
and treatment-related harms. The adaptive approach was worse
calibrated in scenarios with strong linear and non-monotonic
deviations compared to the linear interaction model and RCS-3.

@@

We used the risk linear predictor to fit the proposed methods under
study for individualizing absolute benefit predictions. All methods
predicted increasing benefits for patients with higher baseline risk
predictions, but the fitted patterns were clearly different. The
adaptive approach selected the model with RCS smoothing with 4
knots. However, for very low baseline risk this model predicted
decreasing benefit with increasing risk may be somewhat too flexible.
The more robust models, the linear interaction model or the model with
RCS smoothing (3 knots), gave very similar benefit predictions,
followed the evolution of the stratified estimates very closely and
may therefore be preferable for use in clinical practice. The linear
interaction model had somewhat lower AIC compared to the model with
RCS smoothing (3 knots), slightly better cross-validated
discrimination (c-for-benefit
`r round(discriminationPerformance[2, ], 3)` vs
`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
`r round(calibrationPerformance[3, ], 4)`).

@@

The linear interaction model displayed very good performance overall
under many of the considered simulation scenarios. Especially in cases
with smaller sample sizes and moderately performing baseline risk
prediction models it had lower RMSE, was better calibrated for benefit
and had better discrimination for benefit, even in scenarios with
strong quadratic deviations. However, in scenarios with true
non-monotonic deviations, the linear interaction model was
outperformed by RCS-3, especially in the presence of true
treatment-related harms. Increasing the sample size or the prediction
model's discrimination favored RCS-3 which had better or very
comparable performance to the linear interaction model, but was more
robust to non-monotonic deviations and the presence of
treatment-related harms.

RCS-4 and RCS-5 proved to be too flexible, as indicated by higher
RMSE, increased variability of discrimination for benefit and worse
calibration of benefit predictions. Even with larger sample sizes and
strong quadratic or non-monotonic deviations from the base case
scenario of constant relative treatment effects, these more flexible
restricted cubic splines did not outperform the simpler RCS-3 These
approaches may only be helpful if we expect more extreme patterns of
heterogeneous treatment effects compared to the quadratic deviations
considered here.

The constant treatment effect model, despite having adequate
performance in the presence of weak treatment effect heterogeneity on
the relative scale, quickly broke down with stronger deviations from
constant relative treatment effects. In these cases, the stratified
approach generally had lower error rates compared to the constant
treatment effect model. Stepwise treatment benefit estimates are very
useful for demonstrating treatment effect heterogeneity--because
estimating treatment effect requires groups of patients rather than
individual patients--but are not helpful for making individualized
absolute benefit predictions.

Increasing the discriminative ability of the risk model--by increasing
the predictor coefficients of the true risk model--reduced RMSE for
all methods.  This increase in discriminative ability translates in
higher variability of predicted risks, which, in turn, allows the
considered methods to better capture absolute treatment benefits. As a
consequence, the increase in discriminative ability of the risk model
also led to higher values of c-for-benefit. Even though risk model
performance is very important for the ability of risk-based methods to
predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective**: To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting**: We simulated RCT data using diverse assumptions for the
 average treatment effect, a baseline prognostic index of risk (PI),
 the shape of its interaction with treatment (none, linear, quadratic
 or non-monotonic) and the magnitude of treatment-related harms. In
 each sample we predicted absolute benefit using: models with a
 constant relative treatment effect; stratification in quarters of the
 PI; models including a linear interaction of treatment with the PI;
 models including an interaction of treatment with a restricted cubic
 spline (RCS) transformation of the PI; an adaptive approach using
 Akaike’s Information Criterion. We evaluated predictive performance
 using root mean squared error and measures of discrimination and
 calibration for benefit. **Results**: The linear-interaction model
 and the RCS-interaction model outperformed the constant treatment
 effect model in many simulation scenarios. The RCS-model was optimal
 when quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. **Conclusion**: An
 interaction between risk and treatment assignment generally improved
 treatment effect predictions. Non-linear RCS interactions should be
 considered for larger sample size.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: Ewout W. Steyerberg
    affiliation: 2
  - name: David van Klaveren
    affiliation: 3
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 3
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

d <- function(x, decimals = 4) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. This two-step approach identified substantial absolute treatment
effect differences between low-risk and high-risk patients in a re-analysis of
32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
level may be accurate, these estimates do not apply to individual patients,
especially for patients with predicted risk at the boundaries of the risk
intervals. Hence, the risk-stratified approach is useful for exploring and
presenting HTE, but is not useful for supporting treatment decisions for
individual patients.

@@

We simulated a typical RCT that is undertaken to compare a binary outcome
(e.g. death) between a group of treated patients in the treatment arm and a
group of untreated patients in the control arm. For each patient we generated 8
baseline covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1,
0.2)$. Treatment was allocated using a 50:50 split. Outcomes for patients in the
control arm were generated from a logistic regression model including all
baseline covariates. In the base scenarios coefficient values were such, that
the AUC of the logistic regression model was $0.75$ and the event rate in the
control arm was $20\%$. Binary outcomes in the control arm were generated from
Bernoulli variables with true probabilities
$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.

Outcomes in the treatment arm were generated using 3 base scenarios: absent
treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and strong
treatment effect (OR = 0.5). We started with simulating outcomes based on true
constant relative treatment effects for the 3 base scenarios. We then simulated
linear, quadratic and non-monotonic deviations from constant treatment effects
using:
$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
constant absolute harm is applied across all treated patients. In this case we
have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,063 (4,250 divided by 4) and 17,000 (4250 multiplied by 4),
respectively. We also evaluated the effect of worse or better discriminative
ability for risk, adjusting the baseline covariate coefficients, such that the
AUC of the regression model in the control arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 648 scenarios
(exact settings in the supplementary material). With these scenarios we were
able to cover the observed treatment effect heterogeneity in 32 large trials as
well as many other potential variations of risk-based treatment effect
[@Kent2016].

@@

All risk-based methods assume that a risk prediction model is available to
assign risk predictions to individual patients. For the simulations we developed
a prediction model internally, using logistic regression including main effects
for all baseline covariates and treatment assignment. Risk predictions for
individual patients were based on treatment assignment to the control arm, that
is setting treatment assignment to 0.

@@

```{r adaptive, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveSelections.csv"))
rmseDistribution <- readr::read_csv(here::here("data/processed/rmseDistribution.csv"))
aucDistribution <- readr::read_csv(here::here("data/processed/discriminationDistribution.csv"))
```
The linear interaction model outperformed all RCS methods in terms of
RMSE in scenarios with true constant relative treatment effect (OR =
0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
quadratic deviations from a constant relative treatment effect
(Figure \ref{fig:rmsebase}; panels A-C). However, with
non-monotonic deviations from a constant relative treatment effect,
the RMSE of the linear interaction model increased substantially, 
especially in the presence of treatment-related harms (Figure 
\ref{fig:rmsebase}; panel D). In these scenarios, RCS-3 outperformed all other
methods in terms of RMSE. The constant treatment effect
approach had overall best performance under true constant treatment
effect settings, but was sensitive to all considered deviations,
resulting in increased RMSE. Finally, the adaptive approach had
comparable performance to the best-performing method in each
scenario. However, in comparison with the best-performing approach, its RMSE was
more variable across the 1000 replications in the scenarios with linear and
non-monotonic deviations, especially when also including moderate or strong
treatment-related harms. This is caused by wrongly selecting the constant
treatment effect model in a substantial proportion of the replications
(Supplement, Table XX).

In the scenario with a true constant relative treatment effects, the adaptive
approach selected the constant effect model for the majority of the replcations, even in
the presence of treatment-related harms
(`r adaptiveSelections %>% filter(scenarioId == 217) %>% pull(treatment) * 100`\%
of the replications without harms and 
`r adaptiveSelections %>% filter(scenarioId == 219) %>% pull(treatment) * 100`\% 
with strong harms). With strong linear or quadratic deviations and increasing
treatment-related harms the adaptive approach increasingly favored the linear
interaction model. With non-monotonic deviations, we observed increasing
selection frequencies for RCS-3 with increasing treatment harms (from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(rcs_3) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(rcs_3) * 100`\%
with strong harms), whereas selection frequencies for the linear interaction
model dropped from
`r adaptiveSelections %>% filter(scenarioId == 397) %>% pull(risk) * 100`\%
with no harms to
`r adaptiveSelections %>% filter(scenarioId == 399) %>% pull(risk) * 100`\%
with strong harms. Finally, in the case of strong linear and non-monotonic
deviations selection frequencies of the constant treatment effect model remained
high despite increasing treatment harms (Supplement, Table XX).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

Increasing the sample size to `r 17000` favored RCS-3 the most, as it
achieved lowest or close to lowest RMSE across all scenarios (Figure
\ref{fig:rmsesamplesize}). Especially in cases of strong quadratic and
non-monotonic deviations RCS-3 had lower RMSE (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms) compared
to the linear interaction approach (median 
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of the strength of treatment-related
harms. Due to the large sample size, the RMSE of the adaptive approach
was even more similar to the best-performing method, and the constant
relative treatment effect model was less often wrongly selected
(***numbers????***).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. Sample size 17,000 rather than 4250 in Figure \\ref{fig:rmsebase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

When we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = `r 4250`). RCS-3 had the lowest RMSE in the case of strong
quadratic or non-monotonic deviations and very comparable performance to
the -- optimal -- linear interaction model in the case of strong
linear deviations (median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). As observed in the base case
scenario the adaptive approach wrongly selected the constant treatment
effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the strong linear and non-monotonic deviation
scenarios without treatment-related harms, respectively), leading to
more variability of the RMSE.

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

In comparison with the true approach, dsicrimination for benefit in
the scenario with a constant relative treatment effect was only
slightly lower for the linear interaction model, but more
substantially lower for the non-linear RCS approaches (Figure
\ref{fig:discrimination}; panel A). With strong linear or quadratic
deviations from a constant relative treatment effect, all methods
discriminated quite similarly (Figure \ref{fig:discrimination}; panels
B-C). In the scenario with non-monotonic deviations, the constant
effect model had much lower discriminative ability compared to
non-linear RCS methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`
for the best-performing RCS-3; Figure \ref{fig:rmseauc}; panel D).
With increasing number of RCS knots, we observed decreasing median
values and increasing variability of the c-for-benefit in all
scenarios. ***With increasing sample size... (Suppl Fig...)***

@@

In terms of calibration for benefit, the constant effects model
outperformed all other models in the scenario with true constant
treatment effects, but was miscalibrated for all deviation scenarios
(Figure \ref{fig:calibration}). The linear interaction model showed
best or close to best calibration across all scenarios and only showed
worse calibration compared to RCS-3 in case of non-monotonic
deviations and treatment-related harms (Figure \ref{fig:calibration};
panel D). The adaptive approach was worse calibrated in scenarios with
strong linear and non-monotonic deviations compared to the linear
interaction model and RCS-3. ***With increasing sample size... (Supp
Fig...)***

@@

We used the risk linear predictor to fit the proposed methods under
study for individualizing absolute benefit predictions. All methods
predicted increasing benefits for patients with higher baseline risk
predictions, but the fitted patterns were clearly different. The
adaptive approach selected the model with RCS smoothing with 4
knots. However, for very low baseline risk the decreasing predicted
benefit with increasing risk may be somewhat too flexible. The more
robust models, the linear interaction model or the model with RCS
smoothing (3 knots), gave very similar benefit predictions, followed
the evolution of the stratified estimates very closely and may
therefore be preferable for use in clinical practice. The linear
interaction model had somewhat lower AIC compared to the model with
RCS smoothing (3 knots), slightly better cross-validated
discrimination (c-for-benefit
`r round(discriminationPerformance[2, ], 3)` vs
`r round(discriminationPerformance[3, ], 3)`) and quite similar cross-validated
calibration (ICI-for benefit `r round(calibrationPerformance[2, ], 4)` vs
`r round(calibrationPerformance[3, ], 4)`).

@@

The linear interaction model and the RCS-3 model both displayed very
good performance under many of the considered simulation scenarios, in
contrast with the constant relative treatment effect model. The linear
interaction model was optimal in cases with smaller sample sizes and
moderately performing baseline risk prediction models, that is, it had
lower RMSE, was better calibrated for benefit and had better
discrimination for benefit, even in scenarios with strong quadratic
deviations. In scenarios with true non-monotonic deviations, the
linear interaction model was outperformed by RCS-3, especially in the
presence of true treatment-related harms. Increasing the sample size
or the prediction model’s discriminative ability favored RCS-3,
especially in scenarios with non-monotonic deviations and in the
presence of treatment-related harms.

RCS-4 and RCS-5 proved to be too flexible in all considered scenarios,
as indicated by higher RMSE, increased variability of discrimination
for benefit and worse calibration of benefit predictions. Even with
larger sample sizes and strong quadratic or non-monotonic deviations
from the base case scenario of constant relative treatment effects,
these more flexible restricted cubic splines did not outperform the
simpler RCS-3 These approaches may only be helpful if we expect more
extreme patterns of heterogeneous treatment effects compared to the
quadratic deviations considered here.

The constant treatment effect model, despite having adequate
performance in the presence of weak treatment effect heterogeneity on
the relative scale, quickly broke down with stronger deviations from
constant relative treatment effects. In these cases, the stratified
approach generally had lower error rates compared to the constant
treatment effect model. Stepwise treatment benefit estimates are very
useful for demonstrating treatment effect heterogeneity–because
estimating treatment effect requires groups of patients rather than
individual patients–but are not helpful for making individualized
absolute benefit predictions.

Increasing the discriminative ability of the risk model–by increasing
the predictor coefficients of the true risk model–reduced RMSE for all
methods. This increase in discriminative ability translates in higher
variability of predicted risks, which, in turn, allows the considered
methods to better capture absolute treatment benefits. As a
consequence, the increase in discriminative ability of the risk model
also led to higher values of c-for-benefit. Even though risk model
performance is very important for the ability of risk-based methods to
predict treatment benefit, prediction model development was outside
the scope of this work and has already been studied extensively
[@vanKlaveren2019; @Burke2014; @Abadie2018].

The adaptive approach had adequate performance, following closely on
average the performance of the “true” model in most
scenarios. However, with smaller sample sizes it tended to “miss” the
treatment-risk interactions and selected simpler models (Supplementary
Table S7). This resulted in increased RMSE variability in these
scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in
the case of smaller sample sizes the simpler linear interaction model
is a safer choice for predicting absolute benefits.
